---
title: "Netflix Movies and TV Shows Data Analysis"
author: "Shreya Solanki - 2505101200049   Bsc IT(Hons)   Semester - 3 Division - G"
date: "15-09-2026"
output:
  html_document:
    theme: readable
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.8, fig.align = "center")
required <- c("dplyr", "ggplot2", "stringr", "tidyr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) stop("Install required packages: ", paste(missing, collapse = ", "))
library(dplyr); library(ggplot2); library(stringr); library(tidyr); library(knitr)
```

# Introduction

This tiny project analyses Netflix catalogue data to understand the distribution of Movies and TV Shows, release-year patterns, ratings, countries, and genres. It demonstrates an end-to-end R workflow: data import, cleaning, statistical exploration, plots, interpretation, and a reproducible conclusion.

## Research questions

1. Does the dataset contain more Movies or TV Shows?
2. Which release years, ratings, countries, and genres occur most frequently?
3. How has the number of titles released changed over time?

# Data importing

Download a Netflix titles CSV file and save it as `netflix_titles.csv` in the same folder as this `.Rmd` file. This report expects the commonly used columns: `show_id`, `type`, `title`, `director`, `cast`, `country`, `date_added`, `release_year`, `rating`, `duration`, `listed_in`, and `description`.

```{r import}
netflix <- read.csv("netflix_titles.csv", stringsAsFactors = FALSE, na.strings = c("", "NA"))
head(netflix)
```

```{r inspect}
str(netflix)
summary(netflix)
```

# Preprocessing and cleaning

This section standardises text values, converts dates, removes duplicate titles, and removes records with invalid release years. Missing director, cast, country, rating, and duration values are replaced by `Unknown` so they can be counted rather than silently excluded.

```{r cleaning}
required_columns <- c("show_id", "type", "title", "director", "cast", "country",
                      "date_added", "release_year", "rating", "duration", "listed_in", "description")
missing_columns <- setdiff(required_columns, names(netflix))
if (length(missing_columns) > 0) {
  stop("The CSV is missing: ", paste(missing_columns, collapse = ", "))
}

netflix_clean <- netflix %>%
  mutate(
    across(where(is.character), ~trimws(.x)),
    date_added = as.Date(date_added, format = "%B %d, %Y"),
    year_added = as.integer(format(date_added, "%Y")),
    release_year = as.integer(release_year),
    type = factor(type),
    rating = ifelse(is.na(rating) | rating == "", "Unknown", rating),
    country = ifelse(is.na(country) | country == "", "Unknown", country),
    listed_in = ifelse(is.na(listed_in) | listed_in == "", "Unknown", listed_in),
    director = ifelse(is.na(director) | director == "", "Unknown", director),
    duration = ifelse(is.na(duration) | duration == "", "Unknown", duration)
  ) %>%
  distinct(show_id, .keep_all = TRUE) %>%
  filter(!is.na(title), title != "", !is.na(release_year), release_year >= 1900, release_year <= 2100)

colSums(is.na(netflix_clean))
kable(head(netflix_clean), caption = "First six cleaned records")
```

# Statistical exploration

```{r overview}
overview <- netflix_clean %>%
  summarise(
    Total_titles = n(),
    Movies = sum(type == "Movie"),
    TV_Shows = sum(type == "TV Show"),
    Earliest_release = min(release_year),
    Latest_release = max(release_year),
    Unique_ratings = n_distinct(rating),
    Unique_countries = n_distinct(country)
  )
kable(overview, caption = "Netflix catalogue overview")
```

```{r type-and-rating-tables}
type_count <- netflix_clean %>% count(type, sort = TRUE)
rating_count <- netflix_clean %>% count(rating, sort = TRUE) %>% slice_head(n = 10)
kable(type_count, caption = "Titles by content type")
kable(rating_count, caption = "Ten most frequent ratings")
```

# Graphs and plots

```{r type-plot}
ggplot(netflix_clean, aes(x = type, fill = type)) +
  geom_bar() +
  labs(title = "Netflix Titles by Content Type", x = "Content type", y = "Number of titles") +
  theme_minimal() + theme(legend.position = "none")
```

```{r release-year-plot}
netflix_clean %>%
  count(release_year, type) %>%
  ggplot(aes(x = release_year, y = n, colour = type)) +
  geom_line(linewidth = 0.8) +
  labs(title = "Titles by Release Year", x = "Release year", y = "Number of titles", colour = "Type") +
  theme_minimal()
```

The line chart shows the catalogue composition by original release year. A fall in the final years may reflect incomplete data or titles not yet fully represented in the source file, rather than an actual production decline.

```{r top-countries}
top_countries <- netflix_clean %>%
  filter(country != "Unknown") %>%
  mutate(country = str_split(country, ",\\s*")) %>%
  tidyr::unnest(country) %>%
  count(country, sort = TRUE) %>%
  slice_head(n = 10)

kable(top_countries, caption = "Top ten production countries")

ggplot(top_countries, aes(x = reorder(country, n), y = n)) +
  geom_col(fill = "firebrick") + coord_flip() +
  labs(title = "Top Ten Production Countries", x = "Country", y = "Number of titles") +
  theme_minimal()
```

```{r top-genres}
top_genres <- netflix_clean %>%
  filter(listed_in != "Unknown") %>%
  mutate(genre = str_split(listed_in, ",\\s*")) %>%
  tidyr::unnest(genre) %>%
  count(genre, sort = TRUE) %>%
  slice_head(n = 10)

kable(top_genres, caption = "Top ten listed genres")

ggplot(top_genres, aes(x = reorder(genre, n), y = n)) +
  geom_col(fill = "darkcyan") + coord_flip() +
  labs(title = "Top Ten Netflix Genres", x = "Genre", y = "Number of titles") +
  theme_minimal()
```

```{r rating-plot}
top_ratings <- netflix_clean %>% count(rating, sort = TRUE) %>% slice_head(n = 10)
ggplot(top_ratings, aes(x = reorder(rating, n), y = n)) +
  geom_col(fill = "goldenrod") + coord_flip() +
  labs(title = "Ten Most Common Content Ratings", x = "Rating", y = "Number of titles") +
  theme_minimal()
```

# Interpretation of results

- The content-type chart directly answers whether Movies or TV Shows dominate the selected Netflix catalogue.
- Country and genre fields can contain more than one value per title; this report separates comma-delimited values before counting, so a co-produced title can contribute to multiple countries or genres.
- The release-year plot describes titles available in the dataset, not Netflix subscriber demand, viewing hours, or revenue.
- Missing metadata is labelled `Unknown`; therefore, the analysis makes data quality visible instead of treating missing information as a real category.

# Findings

- The overview table reports the total number of titles and the split between Movies and TV Shows.
- The release-year chart highlights periods with more titles represented in the catalogue.
- The country and genre charts identify the most frequently represented production locations and categories.
- Results depend on the date, country, and source of the downloaded Netflix dataset; the catalogue changes over time.

# Conclusion

This project successfully applies data import, preprocessing, descriptive analysis, and data visualisation to a Netflix catalogue dataset. The analysis provides a useful overview of content type, release periods, ratings, production countries, and genres. A future extension could analyse the relationship between content rating and genre, compare Movies with TV Shows by country, or join the data with IMDb/TMDb ratings where permitted.

# Reproducibility checklist

- Put `netflix_titles.csv` and this `.Rmd` file in the same folder.
- Install packages with `install.packages(c("rmarkdown", "knitr", "dplyr", "ggplot2", "stringr", "tidyr"))`.
- Open the `.Rmd` file in RStudio and select **Knit > Knit to HTML**.
- Verify that the HTML report is created and every table and plot is visible before submission.
