---
title: "Netflix Movies and TV Shows Data Analysis"
author: "Shreya Solanki - 2505101200049   Bsc IT(Hons)   Semester - 3 Division - G"
date: "15-09-2026"
output:
  pdf_document: default
  html_document:
    theme: readable
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 4.8, fig.align = "center")
required <- c("dplyr", "ggplot2", "stringr", "tidyr")
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing) > 0) stop("Install required packages: ", paste(missing, collapse = ", "))
library(dplyr); library(ggplot2); library(stringr); library(tidyr); library(knitr)
```

# Introduction

This tiny project analyses Netflix catalogue data to understand the distribution of Movies and TV Shows, release-year patterns, ratings, countries, and genres. It demonstrates an end-to-end R workflow: data import, cleaning, statistical exploration, plots, interpretation, and a reproducible conclusion.

# Project Definition

The Netflix Movies and TV Shows Data Analysis project is a data analysis project developed using the R programming language. The project explores a Netflix catalogue dataset to understand the distribution of Movies and TV Shows and examine patterns related to release years, ratings, countries, and genres.

The project includes data importing, preprocessing, statistical analysis, data visualization, interpretation of results, and a reproducible conclusion.

# Objectives

The main objectives of this project are:

1. To analyze the Netflix Movies and TV Shows dataset using R.
2. To identify the distribution of Movies and TV Shows.
3. To examine release-year patterns.
4. To analyze the content ratings.
5. To identify frequently represented production countries.
6. To identify frequently occurring genres.
7. To visualize the results using graphs.
8. To interpret the findings from the analysis.

# Dataset Description

The dataset contains information about Netflix Movies and TV Shows. The main attributes used in this analysis are:

| Attribute | Description |
|---|---|
| `show_id` | Unique identifier of the title |
| `type` | Indicates whether the title is a Movie or TV Show |
| `title` | Name of the title |
| `director` | Director of the title |
| `cast` | Cast information |
| `country` | Production country or countries |
| `date_added` | Date the title was added |
| `release_year` | Original release year |
| `rating` | Content rating |
| `duration` | Movie duration or number of seasons |
| `listed_in` | Genres or categories |
| `description` | Description of the title |

# Tools and Technologies

- R
- RStudio
- R Markdown
- dplyr
- ggplot2
- stringr
- tidyr
- knitr

# Data importing

Download a Netflix titles CSV file and save it as `netflix_titles.csv` in the same folder as this `.Rmd` file. This report expects the commonly used columns: `show_id`, `type`, `title`, `director`, `cast`, `country`, `date_added`, `release_year`, `rating`, `duration`, `listed_in`, and `description`.

```{r import}
netflix <- read.csv("netflix_titles.csv", stringsAsFactors = FALSE, na.strings = c("", "NA"))
head(netflix)
```

```{r inspect}
str(netflix)
summary(netflix)
```

# Preprocessing and cleaning

This section standardises text values, converts dates, removes duplicate titles, and removes records with invalid release years. Missing director, cast, country, rating, and duration values are replaced by `Unknown` so they can be counted rather than silently excluded.

```{r cleaning}
required_columns <- c("show_id", "type", "title", "director", "cast", "country",
                      "date_added", "release_year", "rating", "duration", "listed_in", "description")
missing_columns <- setdiff(required_columns, names(netflix))
if (length(missing_columns) > 0) {
  stop("The CSV is missing: ", paste(missing_columns, collapse = ", "))
}

netflix_clean <- netflix %>%
  mutate(
    across(where(is.character), ~trimws(.x)),
    date_added = as.Date(date_added, format = "%B %d, %Y"),
    year_added = as.integer(format(date_added, "%Y")),
    release_year = as.integer(release_year),
    type = factor(type),
    rating = ifelse(is.na(rating) | rating == "", "Unknown", rating),
    country = ifelse(is.na(country) | country == "", "Unknown", country),
    listed_in = ifelse(is.na(listed_in) | listed_in == "", "Unknown", listed_in),
    director = ifelse(is.na(director) | director == "", "Unknown", director),
    duration = ifelse(is.na(duration) | duration == "", "Unknown", duration)
  ) %>%
  distinct(show_id, .keep_all = TRUE) %>%
  filter(!is.na(title), title != "", !is.na(release_year), release_year >= 1900, release_year <= 2100)

colSums(is.na(netflix_clean))
cleaned_preview <- netflix_clean %>%
  select(show_id, type, title, country, release_year, rating)

kable(
  head(cleaned_preview),
  caption = "First six cleaned records"
)
```

# Statistical exploration

```{r overview}
overview <- netflix_clean %>%
  summarise(
    Total_titles = n(),
    Movies = sum(type == "Movie"),
    TV_Shows = sum(type == "TV Show"),
    Earliest_release = min(release_year),
    Latest_release = max(release_year),
    Unique_ratings = n_distinct(rating),
    Unique_countries = n_distinct(country)
  )
kable(overview, caption = "Netflix catalogue overview")
```

```{r type-and-rating-tables}
type_count <- netflix_clean %>% count(type, sort = TRUE)
rating_count <- netflix_clean %>% count(rating, sort = TRUE) %>% slice_head(n = 10)
kable(type_count, caption = "Titles by content type")
kable(rating_count, caption = "Ten most frequent ratings")
```

# Graphs and plots

```{r type-plot}
ggplot(netflix_clean, aes(x = type, fill = type)) +
  geom_bar() +
  labs(title = "Netflix Titles by Content Type", x = "Content type", y = "Number of titles") +
  theme_minimal() + theme(legend.position = "none")
```

```{r release-year-plot}
release_counts <- netflix_clean %>%
  count(release_year)

ggplot(release_counts, aes(x = release_year, y = n)) +
  geom_col() +
  labs(
    title = "Number of Titles by Release Year",
    x = "Release year",
    y = "Number of titles"
  ) +
  scale_y_continuous(
    breaks = seq(0, max(release_counts$n), by = 1)
  ) +
  theme_minimal()
```

The bar chart shows the distribution of titles by original release year. A lower number of titles in some years may reflect the limited size of the dataset rather than an actual production decline.

```{r top-countries}
top_countries <- netflix_clean %>%
  filter(country != "Unknown") %>%
  mutate(country = str_split(country, ",\\s*")) %>%
  tidyr::unnest(country) %>%
  count(country, sort = TRUE) %>%
  slice_head(n = 10)

kable(
  top_countries,
  caption = "Production Countries in the Dataset"
)

ggplot(top_countries, aes(x = reorder(country, n), y = n)) +
  geom_col(fill = "firebrick") +
  coord_flip() +
  labs(
    title = "Production Countries in the Dataset",
    x = "Country",
    y = "Number of titles"
  ) +
  theme_minimal()
```

```{r top-genres}
top_genres <- netflix_clean %>%
  filter(listed_in != "Unknown") %>%
  mutate(genre = str_split(listed_in, ",\\s*")) %>%
  tidyr::unnest(genre) %>%
  count(genre, sort = TRUE) %>%
  slice_head(n = 10)

kable(top_genres, caption = "Most Frequently Listed Genres")

ggplot(top_genres, aes(x = reorder(genre, n), y = n)) +
  geom_col(fill = "darkcyan") + coord_flip() +
  labs(title = "Most Frequently Listed Genres", x = "Genre", y = "Number of titles") +
  theme_minimal()
```

```{r rating-plot}
top_ratings <- netflix_clean %>% count(rating, sort = TRUE) %>% slice_head(n = 10)
ggplot(top_ratings, aes(x = reorder(rating, n), y = n)) +
  geom_col(fill = "goldenrod") + coord_flip() +
  labs(title = "Ten Most Common Content Ratings", x = "Rating", y = "Number of titles") +
  theme_minimal()
```

# Key Observations

- The dataset contains 10 titles, including 6 Movies and 4 TV Shows.
- Movies represent the larger content category in the selected dataset.
- The release years represented in the dataset range from 2018 to 2024.
- TV-14 is the most frequently occurring content rating.
- India is the most frequently represented production country after separating multiple-country entries.
- Dramas is the most frequently listed genre after separating multiple-genre entries.

# Interpretation of results

- The content-type chart directly answers whether Movies or TV Shows dominate the selected Netflix catalogue.
- Country and genre fields can contain more than one value per title; this report separates comma-delimited values before counting, so a co-produced title can contribute to multiple countries or genres.
- The release-year plot describes titles available in the dataset, not Netflix subscriber demand, viewing hours, or revenue.
- Missing metadata is labelled `Unknown`; therefore, the analysis makes data quality visible instead of treating missing information as a real category.

# Findings

- The overview table reports the total number of titles and the split between Movies and TV Shows.
- The release-year chart highlights periods with more titles represented in the catalogue.
- The country and genre charts identify the most frequently represented production locations and categories.
- Results depend on the date, country, and source of the downloaded Netflix dataset; the catalogue changes over time.

# Conclusion

This project successfully applies data import, preprocessing, descriptive analysis, and data visualisation to a Netflix catalogue dataset. The analysis provides a useful overview of content type, release periods, ratings, production countries, and genres. A future extension could analyse the relationship between content rating and genre, compare Movies with TV Shows by country, or join the data with IMDb/TMDb ratings where permitted.

# Reproducibility checklist

## Reproducibility Checklist

1. Place `netflix_titles.csv` and the `.Rmd` file in the same folder.
2. Install the required R packages:
   `rmarkdown`, `knitr`, `dplyr`, `ggplot2`, `stringr`, and `tidyr`.
3. Open the `.Rmd` file in RStudio.
4. Select **Knit → Knit to PDF**.
5. Verify that all tables, graphs, and sections are displayed correctly.